\vspace{-1.5cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2015-09-01 write_up/cricket.tex
More writing

The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging. The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours. And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{type}). But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%. The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss. That can best be ascribed to choosing badly.

@@

But type of matches cover only major source of variation and theorizing about the advantage granted by the toss. It is often claimed that the toss is more crucial in day and night matches. Due to dew -- it is thought to make bowling hard, and lower visibility of the white ball under lights -- it allegedly makes catching hard, it is thought that the team that fields second is disadvantaged. It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{dn}). In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss is at least 3\%, and in case of one-day internationals, nearly 8\%.

@@

@@

@@

>>>>

The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively. And there is good dose of common sense behind the conventional wisdom. Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging. The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours. And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}). But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%. The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss. That can best be ascribed to choosing badly.

@@

But type of matches cover only major source of variation and theorizing about the advantage granted by the toss. It is often claimed that the toss is more crucial in day and night matches. Due to dew -- it is thought to make bowling hard, and lower visibility of the white ball under lights -- it allegedly makes catching hard, it is thought that the team that fields second is disadvantaged. It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{fig:dn}). In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss is at least 3\%, and in case of one-day internationals, nearly 8\%.

@@

In limited over matches sometimes weather intervenes significantly after the match has already begun. After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis. We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth Lewis and matches that don't. We find scant differences except in the case of international Twenty20s (see Figure ~\ref{fig:dl}).

@@

Lastly, winning a toss ought to matter the most when two closely matched teams are playing. What good would winning the toss do if you are matched with a team you have no chance beating. To study this, we collected ranking data. The ranking data are only available for international teams. And for reasons to do with convenience, we limited ourselves to collecting data on one-day internationals and tests between the earliest starting point and 2013. (These ranking data are updated monthly.) We measure how closely matched the teams are by simply subtracting the ranking points of team that lost the toss from the team that won it. The results are expected but new. As Figure ~\ref{fig:ranks}, there is a sharp curve around 0. When closely matched teams win, the winning ratio is strongly affected by who wins the toss.

@@

This isn't a comprehensive inventory of analyses that one could do on the topic. But neither is it meant to be. There are some simple questions that still unanswered. For one, how does the toss advantage vary over time? Have teams become smarter? How the advantage vary by country -- are some countries better than others? Other more nuanced but commonly thought questions also remain. For one, students of the game suspect that the advantage of winning a toss in early English season is especially great. We propose to investigate these questions in a follow-up piece.
